---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file, then run rmarkdown::render("README.Rmd"). -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# neuralsbi

<!-- badges: start -->
[![R-CMD-check](https://github.com/pedroliman/neuralsbi/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/pedroliman/neuralsbi/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/pedroliman/neuralsbi/actions/workflows/pkgdown.yaml/badge.svg)](https://pedroliman.github.io/neuralsbi/)
<!-- badges: end -->

`neuralsbi` is an R-native package for [Neural Simulation-based inference](https://simulation-based-inference.org).

Neural estimators are implemented directly in R on the [`torch`](https://torch.mlverse.org/) R package.

## Installation

```r
# install.packages("remotes")
remotes::install_github("pedroliman/neuralsbi")

# the neural back end (once)
install.packages("torch")
torch::install_torch()
```

## Usage

Simulation-based inference recovers parameters from a model you can simulate but whose likelihood you would rather not write down. To make that concrete, here is a plain linear regression, `y = X beta + noise`. We know the answer (ordinary least squares), so we can check that the posterior lands on the truth.

```{r usage, message = FALSE, warning = FALSE}
library(neuralsbi)
set.seed(1)

# A linear regression: intercept + two covariates, 100 observations.
N <- 100
X <- cbind(1, matrix(rnorm(N * 2), N, 2))   # design matrix
beta_true <- c(1, -2, 0.5)                  # the coefficients to recover
sigma <- 0.5                                # noise standard deviation

# Simulator: given coefficients, simulate a data set and summarise it by the
# least-squares fit. Those fitted coefficients are the "data" the posterior sees.
fit_ols   <- function(y) lm.fit(X, y)$coefficients
simulator <- function(theta) {
  t(apply(theta, 1, function(beta) fit_ols(X %*% beta + rnorm(N, sd = sigma))))
}

# Prior over the three coefficients, then train the neural posterior.
prior <- prior_uniform(low = rep(-3, 3), high = rep(3, 3))
fit   <- npe(prior, simulator, n_simulations = 3000)

# Observe one data set generated from beta_true, then infer the coefficients.
y_obs <- X %*% beta_true + rnorm(N, sd = sigma)
post  <- posterior(fit, x_obs = fit_ols(y_obs))
draws <- sample(post, 10000)
```

The posterior mean recovers the ground-truth coefficients:

```{r recovery}
rbind(truth = beta_true, posterior_mean = colMeans(draws))
```

```{r pairplot, fig.alt = "Pairwise posterior over the three regression coefficients."}
pairplot(draws)     # joint and marginal views of the posterior
```

The same posterior supports point estimates and calibration checks:

```{r diagnostics, message = FALSE, results = "hide"}
map_estimate(post)     # posterior mode
sbc(fit, simulator)    # simulation-based calibration
```

If you're interested in sbi in other languages or functionality not available here, see the [awesome neural SBI repo](https://github.com/smsharma/awesome-neural-sbi); there are some good implementations in python and in Julia.

## Learn more

The [package website](https://pedroliman.github.io/neuralsbi/) has four vignettes that build on each other:

1. [Getting started](https://pedroliman.github.io/neuralsbi/articles/neuralsbi.html) — the core prior/simulator/posterior workflow.
2. [Choosing a density estimator](https://pedroliman.github.io/neuralsbi/articles/density-estimators.html) — MDN, MAF, NSF, and the torch-free baseline.
3. [Checking the posterior](https://pedroliman.github.io/neuralsbi/articles/diagnostics.html) — calibration and predictive diagnostics.
4. [Case study: an SIR epidemic model](https://pedroliman.github.io/neuralsbi/articles/sir-epidemic.html) — the full Bayesian workflow on an applied problem.

## License

MIT
