---
title: "Choosing a density estimator"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a density estimator}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`npe()` approximates the posterior with a *conditional density estimator*: a
model of `p(theta | x)` trained on simulated `(theta, x)` pairs. The
`density_estimator` argument picks which one, and the choice matters most when
the posterior is non-Gaussian. This vignette walks through the four options and
shows where each earns its place.

## The four estimators

- `"linear_gaussian"` — a closed-form conditional Gaussian fit by linear
  regression. No neural network and no `torch`. It is *exact* when the true
  posterior is Gaussian in the parameters, which makes it both a fast baseline
  and a sanity check for the neural estimators.
- `"mdn"` (the default) — a Mixture Density Network: a small MLP that maps `x`
  to the weights, means, and covariances of a Gaussian mixture. Mixtures handle
  multimodal and skewed posteriors, and training is quick.
- `"maf"` — a Masked Autoregressive Flow (Papamakarios et al., 2017): a stack
  of invertible autoregressive transforms with exact densities. More flexible
  than a mixture when the posterior has curved, sharp structure.
- `"nsf"` — a Neural Spline Flow (Durkan et al., 2019): an autoregressive flow
  whose transforms are monotonic rational-quadratic splines. The most
  expressive option per layer, at somewhat higher training cost.

All three neural estimators need the `torch` back end
(`install.packages("torch"); torch::install_torch()`) and share one training
loop, so the training arguments to `npe()` — `max_epochs`, `patience`,
`n_restarts`, and friends — mean the same thing regardless of which you pick.

## A posterior no Gaussian can fit

The **two-moons** benchmark task has a bimodal, crescent-shaped posterior. It
ships with the package:

```{r}
library(neuralsbi)

task <- task_two_moons()
task
#> <nsbi_task> two_moons: 2 parameters -> 2 data dims

x_obs <- c(0, 0)
```

The linear-Gaussian estimator is structurally unable to represent two modes —
it collapses them into one wide blob:

```{r}
fit_lg <- npe(task$prior, task$simulator, n_simulations = 10000,
              density_estimator = "linear_gaussian", seed = 1)
draws_lg <- sample(posterior(fit_lg, x_obs = x_obs), 5000)
pairplot(draws_lg)
```

The MDN, with its mixture components, recovers both moons:

```{r}
fit_mdn <- npe(task$prior, task$simulator, n_simulations = 10000,
               density_estimator = "mdn", n_components = 10, seed = 1)
draws_mdn <- sample(posterior(fit_mdn, x_obs = x_obs), 5000)
pairplot(draws_mdn)
```

And a spline flow captures the sharp crescent edges most cleanly:

```{r}
fit_nsf <- npe(task$prior, task$simulator, n_simulations = 10000,
               density_estimator = "nsf", seed = 1)
draws_nsf <- sample(posterior(fit_nsf, x_obs = x_obs), 5000)
pairplot(draws_nsf)
```

## Comparing estimators quantitatively

A classifier two-sample test (`c2st()`) measures how distinguishable two sets
of samples are: 0.5 means a classifier cannot tell them apart, 1.0 means it
always can. Comparing each estimator's draws against the flow's shows how much
posterior structure the simpler models miss:

```{r}
c2st(draws_mdn, draws_nsf)$accuracy   # close to 0.5: MDN and NSF agree
c2st(draws_lg,  draws_nsf)$accuracy   # well above 0.5: the Gaussian misses modes
```

When an analytic reference posterior exists — as for
`task_gaussian_linear()` — the same test scores an estimator against the exact
answer; that is how the package's own accuracy tests work.

## Capacity knobs

Each estimator exposes a small number of capacity controls through `npe()`:

| Estimator | Arguments | Default |
|---|---|---|
| `"mdn"` | `n_components`, `hidden` | 5 components, two hidden layers of 50 units |
| `"maf"`, `"nsf"` | `n_transforms`, `hidden` | 5 transforms |
| `"linear_gaussian"` | none | — |

Raising capacity helps only if training data supports it; with a few thousand
simulations the defaults are usually right. When results look underfit,
increasing `n_simulations` typically buys more than increasing network size.

## Practical guidance

- Start with the default `"mdn"`. It trains fast and handles multimodality.
- If the posterior has sharp or strongly curved structure the MDN smooths
  over, switch to `"nsf"` (or `"maf"` for a lighter flow).
- Use `"linear_gaussian"` when `torch` is unavailable, when you want a
  near-instant baseline, or when the model really is linear-Gaussian — then it
  is exact.
- Whatever you choose, check it. `vignette("diagnostics")` shows how to verify
  a fitted posterior with calibration and predictive checks.
