---
title: "Case study: inferring epidemic parameters (SIR)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: inferring epidemic parameters (SIR)}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette works an applied, likelihood-free problem end to end: recovering
the transmission and recovery rates of an epidemic from a noisy incidence
curve. There is no tractable likelihood — the data come from a stochastic
compartmental simulator — which is exactly where simulation-based inference
earns its keep.

## The model

The classic **SIR** model splits a population of size `N` into Susceptible,
Infected, and Recovered compartments. Two rates govern the dynamics: the
contact rate \(\beta\) (how fast the infection spreads) and the recovery rate
\(\gamma\) (how fast infected individuals recover). `neuralsbi` ships this as a
built-in benchmark task.

```{r}
library(neuralsbi)

task <- task_sir()          # prior + simulator for the SIR model
task
#> <nsbi_task> sir: 2 parameters -> 10 data dims
```

The task bundles a `prior` (log-normal on \(\beta, \gamma\)) and a
`simulator` that solves the SIR dynamics and returns the observed infected
fraction at 10 time points.

## Fit an amortized posterior

We simulate from the prior and train a neural posterior estimator. A Mixture
Density Network is a good default here; the posterior is smooth and unimodal.

```{r}
fit <- npe(task$prior, task$simulator,
           n_simulations = 10000,
           density_estimator = "mdn",
           seed = 1)
```

Training is *amortized*: this one fit can be conditioned on any incidence curve
without re-simulating.

## Condition on an observation

Suppose we observe an outbreak generated by \(\beta = 0.4,\ \gamma = 0.125\)
(basic reproduction number \(R_0 = \beta/\gamma = 3.2\)).

```{r}
theta_true <- c(beta = 0.4, gamma = 0.125)
x_obs <- task$simulator(matrix(theta_true, nrow = 1))

post <- posterior(fit, x_obs = x_obs)
summary(post, n = 5000)
#>   parameter  mean     sd   q2.5   q50  q97.5
#> 1    theta1 0.40.. 0.0... ...    ...   ...
#> 2    theta2 0.12.. 0.0... ...    ...   ...

draws <- sample(post, 10000)
pairplot(draws, truth = theta_true)
```

The posterior concentrates around the true rates, and — importantly — reports
its own uncertainty.

## Is the posterior calibrated?

A posterior is only trustworthy if it is *calibrated*. We check with
Simulation-Based Calibration and an expected-coverage plot, neither of which
needs a reference posterior.

```{r}
res <- sbc(fit, task$simulator, n_sbc = 300, n_posterior_samples = 1000,
           seed = 2)
res                     # per-parameter uniformity p-values (large = good)

plot_sbc(res, param = 1)   # rank histogram: flat = calibrated
plot_coverage(res)         # empirical vs nominal coverage: on the diagonal = good
```

If the rank histograms are flat and the coverage curve hugs the diagonal, the
posterior's credible intervals mean what they say: a 90% interval contains the
truth about 90% of the time.

## Posterior predictive check

Finally, push posterior draws back through the simulator and compare the
predicted incidence curves to the observation.

```{r}
pp <- posterior_predictive(post, task$simulator, n = 200)
matplot(t(pp), type = "l", col = adjustcolor("grey", 0.3),
        xlab = "observation time", ylab = "infected fraction")
lines(as.numeric(x_obs), col = "firebrick", lwd = 2)
```

The observed curve should sit comfortably within the cloud of predictive draws.
A systematic mismatch would flag model misspecification — a signal no
point estimate can give you.

## Spending simulations where they matter: sequential NPE

Amortization is wasteful when you care about a *single* observation: most
prior draws produce epidemics nothing like the one observed, and the network
spends capacity learning posteriors you will never query. Sequential NPE
(`npe_sequential()`, using truncated proposals) alternates training with
simulation, restricting each new round of simulations to the region the
current posterior considers plausible.

```{r}
fit_seq <- npe_sequential(task$prior, task$simulator, x_obs = x_obs,
                          n_rounds = 3, n_simulations = 3000,
                          density_estimator = "mdn", seed = 3)
fit_seq                    # per-round budgets and acceptance rates

post_seq  <- posterior(fit_seq, x_obs = x_obs)
draws_seq <- sample(post_seq, 10000)
pairplot(draws_seq, truth = theta_true)
```

With the same total simulation budget (here 9,000 vs. 10,000), the sequential
fit typically yields a tighter posterior around this particular outbreak. The
trade-off: the result is specific to `x_obs`, so conditioning on a different
incidence curve means refitting.

## Where to go next

This case study exercised the whole workflow: prior, simulator, amortized
training, conditioning, calibration checks, predictive checks, and a
sequential refinement. The earlier vignettes cover each stage in more depth —
`vignette("neuralsbi")` for the core API,
`vignette("density-estimators")` for when to reach for `"maf"` or `"nsf"`
instead of the MDN, and `vignette("diagnostics")` for the full checking
toolbox.
